---
title: "Sample Size"
---

# Sample Size to Study Rare Events

The hypothetical example is a study comparing two drugs with 150 participants in each arm. The primary outcome is whether a seizure develops after traumatic brain injury.  The estimated incidence of seizure is 2%

Several hypothetical possible outcomes of the study are then arbitrarily generated.  For argument sake, the patients in the Drug A treatment arm experience 3 seizures.  Various *synthetic* data sets are then generated with between 0 and 10 seizures in the Drug B group.


## Scenario 1

In this scenario, Drug A is 100% effective and Drug B has no effect.  The Drug A arm has zero seizures and the Drug B arm has 3 seizures:

```{r,echo=F}
seizure <- c(0, 3)
no_seizure <- c(150-seizure[1],150-seizure[2])

study.data <- data.frame(
  seizure,
  no_seizure,
  stringsAsFactors = TRUE
)
row.names(study.data) <- c("A","B")
print(study.data)

fisher.test(study.data)

```
Fisher's exact test shows no statistical difference between arms.

## Scenario 2

In this scenario, Drug A is 100% effective and Drug B decreases the seizure threshold, resulting in more seizures than expected 5/150 = 3.3%

```{r,echo=F}
seizure <- c(0, 5)
no_seizure <- c(150-seizure[1],150-seizure[2])

study.data <- data.frame(
  seizure,
  no_seizure,
  stringsAsFactors = TRUE
)
row.names(study.data) <- c("A","B")
print(study.data)

fisher.test(study.data)

```

In this scenario, Fisher's exact test still does not demonstrate a statistically significant differene between groups.


## Scenario 3

In this scenario, Drug A is 100% effective and Drug B decreases the seizure threshold, doubling the number of seizures:

```{r,echo=F}
seizure <- c(0, 6)
no_seizure <- c(150-seizure[1],150-seizure[2])

study.data <- data.frame(
  seizure,
  no_seizure,
  stringsAsFactors = TRUE
)
row.names(study.data) <- c("A","B")
print(study.data)

fisher.test(study.data)

```
In this case, a study with 150 subjects per arm could detect a difference where one drug is 100% effective and the other drug doublesY
the baseline seizure rate.  

## Power Calculation

Using the pwr package from R, in order to detect a difference between groups where one group has an event rate of 1% and the other group has an event rate of 2% (assuming 80% power and 0.05 significance):

```{r}
library (pwr)
power12 <-pwr.2p.test(h = ES.h(p1 = 0.01, p2 = 0.02), sig.level = 0.05, power = .80)
power12

```

These calculations suggest a study size of 2,253 patients per arm would be required.

## Conclusion

A trial with 150 subject per arm and a baseline event rate of 2% is not powered to detect a difference between arms except with the most wildly optimistic of assumptions regarding the differences between arms.



```{r, include=F}


power2 <-pwr.2p.test(h = ES.h(p1 = 0.01, p2 = 0.02),
                     n = 150,
                     sig.level = 0.05)



p1 <- seq(0.01, .20, 0.01)
power1 <-pwr.2p.test(h = ES.h(p1 = p1, p2 = 0.02),
                     n = 150,
                     sig.level = 0.05)
powerchange <- data.frame(p1, power = power1$power * 100)

#cat(dev.off())
plot(powerchange$p1, 
     powerchange$power, 
     type = "b", 
     xlab = "Proportion of Responders in Treatment A", 
     ylab = "Power (%)")




```
